Approximating the end of nested sampling

### 2023-07-21 paper/paper.tex
Added sketch of notes on maximisation observations

\documentclass[usenatbib]{mnras}
\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage{bm}
\usepackage{savesym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{relsize}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{afterpage}
\usepackage{txfonts}
\usepackage{pgf}

@@

>>>>

\documentclass[usenatbib]{mnras}
\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage{bm}
\usepackage{savesym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{relsize}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{afterpage}
\usepackage{txfonts}
\usepackage{pgf}
\usepackage{cleveref}

@@

\section{The anatomy of a Gaussian}

In general, it is well known (to those who know such things) that the anatomy of a Gaussian says that
\begin{equation}
    \boxed{\log\mathcal{L}_\mathrm{max} = \langle\log\mathcal{L}\rangle_\mathcal{P} + \frac{d}{2}},  \qquad \boxed{\mathrm{var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}},
\end{equation}
namely the posterior average of the log likelihood is $d/2$ less than the maximum, and the variance of the loglikelihood over the posterior is $\frac{d}{2}$. Whilst this is exact for gaussians, this provides a good rule of thumb for other distributions.

For example, one may use this to explain why MCMC sampling is a very bad way to maximise, since even in the limit of a very large number of samples from the posterior $n$, the largest of these will be within a few $\sqrt{\frac{d}{2}}$ standard deviations of the mean, and therefore very far away from the maximum $\frac{d}{2}$ away.\footnote{TODO: can we do some analytics on the MCMC eaxmple as well?}

The same is true for nested sampling, although since it is not a posterior sampler, but a likelihood upward scanner, the argument is a little more subtle, which we work through below.

We will work with a Gaussian loglikelihood, lengthscale $\sigma$, defined in terms of prior volumes $0<X<1$
\begin{equation}
    \log\mathcal{L}
    =\log\mathcal{L}_\mathrm{max} - \frac{1}{2\sigma^2}X^{2/d}
    =\log\mathcal{L}_\mathrm{max} - \frac{1}{2\sigma^2}\exp\left(\frac{2}{d}\log X\right).
    \label{eq:LX}
\end{equation}
We will assume the run has progressed far enough so that prior effects may be neglected, and that $\sigma\ll 1$ is sufficiently small that for the arguments below edge effects may be neglected.

Note that from the Gaussian profile we can see that as $\log X\to-\infty$ at a steady rate of compression, we approach the maximum loglikelihood exponentially slowly, with an decay rate of $\frac{d}{2}$

Taking $X \sim e^{-i/n}$ we see this as
\begin{equation}
    \log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \exp\left(-\frac{i-i_0}{nd/2}\right), \qquad i_0 = -\frac{nd}{2}\log 2\sigma^2
\end{equation}
in principle this converges very quickly, although never quite reaches the maximum.

\section*{The anatomy of a gaussian}
\begin{equation}
    P(\log\mathcal{L}) = \frac{1}{\Gamma(\frac{d}{2})}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})\sim \chi^2_{d} $

\section*{Observation 1: the largest likelihood live point}
Assume we are at an iteration such that $\mathcal{L}_*$ is the current likelihood contour (i.e. the likelihood of the dead point we have just deleted), and that we have $n$ live pointsdistributed uniformly in prior volume beneath this. 
Since the the live points are distributed in prior volume up to $X_*$, $P(X) = [0<X<X_*]$,
Inverting \cref{eq:LX} we may show that the distribution of loglikelihoods is
\begin{equation}
    P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}}[0<\log\mathcal{L}<\log\mathcal{L}_*].
    \label{eq:PL}
\end{equation}
It is helpful at this stage therefore to define a parameter
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*} \qquad\Leftrightarrow\qquad \log\mathcal{L} = y\log\mathcal{L}_\mathrm{max} + (1-y)\log\mathcal{L}_*
    \label{eq:y}
\end{equation}
as a normalised likelihood with $y=0$ corresponding to $\mathcal{L}=\mathcal{L}_*$ and $y=1$ to $\mathcal{L}=\mathcal{L}_\mathrm{max}$, which has distribution
\begin{equation}
    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1}[0<y<1].
    \label{eq:Py}
\end{equation}
We now seek the distribution of the largest likelihood $\log{\mathcal{L}}_\mathrm{max}^\mathrm{live}$ (or equivalently the largest ${y}_\mathrm{max}^\mathrm{live}$) in the set of $n$ live points. We first integrate to find the CDF
\begin{equation}
    P(y<Y) = 1-(1-Y)^{\frac{d}{2}}[0<Y<1].
    \label{eq:Py}
\end{equation}
note that to find the largest of $n$ independent variables we simply multiply 
\begin{equation}
    P(y_1,y_2,\ldots<Y) = (1-(1-Y)^{\frac{d}{2}})^n[0<Y<1].
    \label{eq:Py}
\end{equation}
and then differentiate with respect to $Y$ to find the PDF for largest likelihood in the live points
\begin{equation}
    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
    \label{eq:Pyhat}
\end{equation}
which may be roughly summarised as:
\begin{equation}
    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
    \label{eq:ymax}
\end{equation}
or in the large $d$, $n$ limit
\begin{align}
    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
    \label{eq:ymaxd}\\
    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
    \label{eq:ymaxdn}
\end{align}
where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.

This shows that in general the live points are nowhere near the maximum loglikelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$. In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away. Our end result is that:
\begin{equation}
    \boxed{
        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - (1-y_\mathrm{max}^\mathrm{live})(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)
    },
\end{equation}
where you may take any of the $y_\mathrm{max}^\mathrm{live}$ expressions above from \cref{eq:ymax,eq:ymaxd,eq:ymaxdn}.

We may also rewrite the distribution \cref{eq:Pyhat} without $y$ to give
\begin{equation}
    \frac{\log\mathcal{L}_\mathrm{max}^\mathrm{live}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
\end{equation}
\begin{equation}
    P(\log\mathcal{L}_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left((\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}-(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
\end{equation}

\subsection*{Observation 2: the termination contour $\mathcal{L}_\mathrm{end}$}

One strategy for terminating nested sampling is to halt when the amount of evidence in the live points is some fraction $f$ of the amount of total evidence (or evidence in the dead points). Providing $f\ll 1$, then this is approximately the same as the fraction of evidence left above likelihood contour $\mathcal{L}_\mathrm{end}$:
\begin{equation}
    f = \frac{\int_0^{X_\mathrm{end}} \mathcal{L} dX}{\int_0^\infty \mathcal{L} dX}.
    \label{eq:fint}
\end{equation}
Note that we have assumed that prior effects are negligible (so $1=\infty$).
Computing this for \cref{eq:LX} we find the answer in terms of lower incomplete gamma functions
\begin{equation}
    f = 1- \frac{\Gamma_{d/2}(\frac{X_\mathrm{end}^{2/d}}{2\sigma^2})}{\Gamma(\frac{d}{2})}.
    \label{eq:f}
\end{equation}
Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
\begin{equation}
    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} f \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d\Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
\end{equation}
We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $f$. This becomes yet more pleasing in the large $d$ limit, since $f^{2/d}\to 1$, we find via a stirling/Stirling approximation:
\begin{equation}
    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
\end{equation}
In the event that we keep $f$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}f^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.

\subsection*{Observation 3: nested sampling as a maximiser}

Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with the large $d,n$ limit of \cref{eq:ymaxdn} we find
\begin{equation}
    \boxed{
        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
    },
\end{equation}
showing that in general the anatomy that nested sampling will finish at a contour $\frac{d}{2e}$ away from the maximum loglikelihood, and the final set of $n$ live points can get you $\frac{\log n}{2e}$ closer, with a chance of getting $\sim\frac{\pi}{\sqrt{6}e}=0.472$ closer still by statistical fluctuation.

\begin{figure*}
\begin{center}
	\includegraphics{Figures/anatomy.pdf}
\end{center}
\caption{}
\label{fig:anatomy}
\end{figure*}
